---
title: "Preoperative Atelectasis"
subtitle: "Part 2: Descriptive characteristics and map"
author: "Javier Mancilla Galindo"
date: "`r Sys.Date()`"
output-dir: /docs/reports
execute: 
  echo: true
  warning: false
toc: true
toc_float: true
format:
  html:
    embed-resources: true
  pdf: 
    documentclass: scrartcl
editor: visual
---

# Setup 

#### Packages used   
```{r}
if (!require("pacman", quietly = TRUE)) {
  install.packages("pacman")
}


pacman::p_load(
  tidyverse, # Used for basic data handling and visualization.
  table1, #Used to create table of descriptive characteristics of sample.
  RColorBrewer, #Color palettes for data visualization. 
  gridExtra, #Used to arrange multiple ggplots in a grid.  
  grid, #Used to arrange multiple ggplots in a grid.
  rnaturalearth, #Used to extract geographical data to create maps. 
  rnaturalearthhires, #Used together with the prior package to create map. 
  sf, #Used together with the prior package to create map. 
  plotly, #Used together with prior two packages to create map.  
  flextable, #Used to export tables.  
  officer  #Used to export tables.
)
```

```{r}
#| include: false

# If you get an error when installing rnaturalearthhires, please run
# this code. If not, ignore this chunk of code.   

if (!require("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

remotes::install_github("ropensci/rnaturalearthhires")

pacman::p_load(rnaturalearthhires)
```


##### Session and package dependencies   
```{r}
#| echo: false

# Credits chunk of code: Alex Bossers, Utrecht University (a.bossers@uu.nl)

# remove clutter
session <- sessionInfo()
session$BLAS <- NULL
session$LAPACK <- NULL
session$loadedOnly <- NULL
# write log file
writeLines(
  capture.output(print(session, locale = FALSE)),
  paste0("sessions/",lubridate::today(), "_session_Part_2.txt")
)

session

rm(session)

```


```{r}
#| include: false

# Create directories for sub-folders 
figfolder <- "../results/output_figures"
tabfolder <- "../results/output_tables"
psfolder <- "../data/processed"
dir.create(figfolder, showWarnings = FALSE)
dir.create(tabfolder, showWarnings = FALSE)
dir.create(psfolder, showWarnings = FALSE)
```


```{r}
#| include: false

# Load dataset  
data <- read.csv("../data/processed/atelectasis_included.csv",
                 na.strings="NA", 
                 row.names = NULL)
# Recode variables 
source("scripts/variable_names.R")

```

# State of residence of participants  

Map generated with the accompanying script ***Map_USA_Canada.R***

This map was built by partly using code adapted from [contribution by cpsievert](https://community.rstudio.com/t/combining-usa-canada-specific-province-with-plotly-plot-geo-or-ggplot2/12375).

```{r}
#| echo: false
#| warning: false

# This may take a few seconds to run, please be patient and ignore warnings. 

source("scripts/Map_USA_Canada.R", local = knitr::knit_global())
fig

rm(canada, usa, fig, fig_add, figS2, table, us_laea)
```


# Distribution of numerical variables    

Distributions were examined with the accompanying sourced function  ***distribution_numerical_variables.R***      
```{r, fig.height = 4, fig.width = 6}
#| include: false

# If plots are not showing and you wish to see them, 
# please change the above to ## include: true ##  

source("scripts/distribution_plots.R")
distribution_plots(data)
```

Near normal distribution:  
-   Age: light tails  
-   height: heavy right tail, 4 outliers right  
-   hb: heavy tails, bilateral outliers    
-   hct: heavy tails, bilateral outliers  
-   leu: near normal, bilateral outliers  
-   neu_absolute: heavy right tail, two right outliers  
-   linf_absolute: heavy right tail, bilateral outliers (more right)     
-   mon_absolute: heavy right tail, bilateral outliers (more right)   
-   platelets: two right outliers   
-   urea: four right outliers  
-   creatinine: three right outliers  


Distribution not normal:  
-   Weight: right-skewed, outliers are verified observations of extreme weight.   
-   BMI: right-skewed, outliers are verified observations of extreme BMI.  
-   spo2_VPO: Left-skewed   
-   neu_percent: left-skewed  
-   linf_percent: right-skewed  
-   glucose: right-skewed  
-   mon_percent: observations around only 5 data points. Will not use this variable, only absolute monocytes will be used.     
-   altitude: distribution not clear as values are quite apart an concentrate around single states with diferring mean altitudes. Will attempt to model a smooth term or categorical term in subsequent analyses.         

Outcome variable:  
-   atelectasis_percent: Zero-inflated. Would be difficult to manage as categorical ordinal due to low number of patients in some categories. Will re-assess alongside subsequent analyses to decide.   


# Characteristics of participants      
Table 1 generated with the accompanying sourced script ***table1_arguments.R***    
```{r}
#| include: false

source("scripts/table1_arguments.R", local = knitr::knit_global())
```

Characteristics of participants are shown for the total sample and by obesity class category as defined by the [World Health Organization](https://www.who.int/europe/news-room/fact-sheets/item/a-healthy-lifestyle---who-recommendations):   

- Class 1, BMI [30,35) kg/m2    
- Class 2, BMI [35,40) kg/m2     
- Class 3, BMI >40 kg/m2     

> Characteristics of participants according to BMI class are shown in **Table 1**.  

## Table 1  
```{r}
#| echo: false

table1
```

NOTE: The **ASA** physical status variable has not been included in analyses since the [updated version of ASA](https://www.asahq.org/standards-and-practice-parameters/statement-on-asa-physical-status-classification-system) consulted in October 2023 classifies obesity (30<BMI<40) as ASA 2 and obesity (BMI ≥40) as ASA 3. The distribution of frequencies of ASA~obesity class in this dataset does not match such definition. This occurred since an outdated version of ASA that did not include obesity was likely used by clinicians when writing the preoperative assessment medical note:   

```{r}
#| echo: false

table(data$ASA,data$type_obesity)
```

```{r}
#| include: false

# Run this chunk if you wish to clear your environment and unload packages.

rm(properties, name, table1, rndr, normal, nonormal, distribution_plots,
   abbreviations, abbreviations_stats,data, figfolder, psfolder, tabfolder)

pacman::p_unload(negate = TRUE)
```
