---
title: "Preoperative Atelectasis"
subtitle: "Part 4: Outcomes"
author: "Javier Mancilla Galindo"
date: "`r Sys.Date()`"
project: 
    output-dir: ../docs/reports
execute: 
  echo: false
  warning: false
toc: true
toc_float: true
format:
  html:
    embed-resources: true
  pdf:
    documentclass: scrartcl
editor: visual
---

# Setup

#### Packages used

```{r}
#| echo: true
if (!require("pacman", quietly = TRUE)) {
  install.packages("pacman")
}


pacman::p_load(
  tidyverse, # Used for basic data handling and visualization.
  table1, #Used to add lables to variables.
  RColorBrewer, #Color palettes for data visualization. 
  gridExtra, #Used to arrange multiple ggplots in a grid.  
  grid, #Used to arrange multiple ggplots in a grid.
  mgcv, #Used to model non-linear relationships with a general additive model.  
  ggmosaic, #Used to create mosaic plots.   
  car, #Used assess distribution of continuous variables (stacked Q-Q plots).
  simpleboot, boot, # Used to calculate mean atelectasis coverage and 
                   # 95%CI through bootstrapping.
  gt #Used to present tables in html format. 
)
```

##### Session and package dependencies

```{r}
# Credits chunk of code: Alex Bossers, Utrecht University (a.bossers@uu.nl)

# remove clutter
session <- sessionInfo()
session$BLAS <- NULL
session$LAPACK <- NULL
session$loadedOnly <- NULL
# write log file
writeLines(
  capture.output(print(session, locale = FALSE)),
  paste0("sessions/",lubridate::today(), "_session_Part_4.txt")
)

session

rm(session)

```

Set seed (for reproducibility of bootstrapping) as the current year 2023:

```{r}
#| echo: true
seed <- 2023 
```

```{r}
#| include: false

# Create directories for sub-folders 
figfolder <- "../results/output_figures"
dir.create(figfolder, showWarnings = FALSE)
```

```{r}
#| include: false

# Load dataset  
data <- read.csv("../data/processed/atelectasis_included.csv",
                 na.strings="NA", 
                 row.names = NULL)
# Recode variables 
source("scripts/variable_names.R")

```

# Outcome variable

```{r}
#| include: false

attach(data)
```

Corroborate that atelectasis(Yes/No) matches atelectasis percent equal or different to 0%:

```{r}
table(atelectasis,atelectasis_percent)
```

Yes, these do match.

## Prevalence of atelectasis

```{r}
frequencies <- table(atelectasis)
percent <- round((prop.table(frequencies)*100),1)
total <- rbind(frequencies,percent)
total
```

Prevalence of atelectasis with 95% confidence interval

```{r}
prev_atelectasis <- prop.test(frequencies, correct=FALSE)
confint <- round((prev_atelectasis$conf.int*100),2)
prev_atelectasis
```

> The prevalence of atelectasis was **`r percent[1]` (95%CI: `r confint`)**.

```{r}
#| include: false
rm(frequencies, percent, prev_atelectasis, total, confint)
```

## Atelectasis - obesity class

Mean expected frequency:

```{r}
mean_exp <- data %>% 
  drop_na(type_obesity, atelectasis) %>% 
  summarize(mean_expected_freq = n()/(nlevels(type_obesity)*nlevels(atelectasis)))

mean_exp
```

Frequencies:

```{r}
frequencies <- table(type_obesity, atelectasis)
frequencies
```

Percentage:

```{r}
round(prop.table(frequencies,1),4)*100
```

Mosaic Plot

```{r, fig.height=3, fig.width=5}
data %>% 
  mutate(atelectasis = fct_relevel(atelectasis, "No", "Yes")) %>%
ggplot() +
  geom_mosaic(
    aes(x = product(atelectasis,type_obesity), 
        fill=atelectasis),
    na.rm = TRUE
    ) +
  scale_fill_manual(values=c("grey95","lightsteelblue4")) +
  labs(
    y = "Atelectasis",
    x = "Obesity class"
  ) +
  theme_mosaic() + 
  theme(axis.text.x=element_text(size=rel(0.8)))
```

```{r}
chi <- chisq.test(frequencies, correct=FALSE)
chi
```

```{r}
#| include: false
rm(mean_exp,frequencies, chi)
```

#### Atelectasis location by obesity class

Mean expected frequency:

```{r}
mean_exp <- data %>% 
  drop_na(type_obesity, atelectasis_location) %>% 
  summarize(mean_expected_freq = n()/(nlevels(type_obesity)*nlevels(atelectasis_location)))

mean_exp
```

Mean expected frequency is greater than 5.0, so chi-squared without continuity correction is adequate.

Frequencies:

```{r}
frequencies <- table(type_obesity, atelectasis_location)
frequencies
```

Percentage:

```{r}
round(prop.table(frequencies,1),4)*100
```

Mosaic Plot

```{r, fig.height=3, fig.width=5}
ggplot(data = data) +
  geom_mosaic(
    aes(x = product(type_obesity,atelectasis_location),
        fill=type_obesity),
    na.rm = TRUE
    ) +
  scale_fill_manual(values=c("seagreen1","seagreen3","seagreen4")) +
  labs(
    y = "Obesity class",
    x = "Atelectasis location"
  ) +
  theme_mosaic() 
```

```{r}
#| warning: false
chi <- chisq.test(frequencies, correct=FALSE)
chi
```

Prevalence of atelectasis with 95% confidence intervals calculated with sourced script ***Prevalence_atelectasis.R***

```{r}
#| include: false

source("scripts/Prevalence_atelectasis.R", local = knitr::knit_global())
```

> The prevalence of atelectasis was greater in higher obesity classes: class 1, n=`r atelectasis$n[3]` (`r atelectasis$prev[3]`%, 95%CI:`r atelectasis$confint[3]`); class 2, n=`r atelectasis$n[5]` (`r atelectasis$prev[5]`%, 95%CI:`r atelectasis$confint[5]`); and class 3, n=`r atelectasis$n[7]` (`r atelectasis$prev[7]`%, 95%CI:`r atelectasis$confint[7]`) (p\<0.001).

> Of those who had atelectasis, the most frequent presentation was unilateral n=`r atelectasis$Unilateral[1]`, compared to bilateral n=`r atelectasis$Bilateral[1]`. When examining this by obesity class, the observed distribution was not significantly different for those with class 1, 2, and 3 obesity categories (n=`r atelectasis$Unilateral[3]`, n=`r atelectasis$Unilateral[5]`, and n=`r atelectasis$Unilateral[7]`, respectively) (p=`r round(chi$p.value,3)`).

```{r}
#| include: false 
rm(mean_exp, frequencies, chi, location, atelectasis_obesity, atelectasis_total, atelectasis_total_location, atelectasis_obesity_location, atelectasis)
```

#### Atelectasis Percent

```{r}
#| include: false 
## Change 'false' for 'true' above to show plot. 

# Assess distribution if assumed to be a numeric variable:   
data %>% 
  mutate(atelectasis_percent = as.numeric(atelectasis_percent)) %>%
  ggplot(aes(x = atelectasis_percent)) +
  geom_histogram(colour = "black") +
  facet_grid(type_obesity ~ .)

```

```{r}
#| include: false 
## Change 'false' for 'true' above to show plot. 

# Distribution excluding 0:  
data %>% 
  mutate(atelectasis_percent = as.numeric(atelectasis_percent)) %>% 
  filter(atelectasis_percent >0) %>% 
  ggplot(aes(x = atelectasis_percent)) +
  geom_histogram(colour = "black") +
  facet_grid(type_obesity ~ .)

```

##### Mean atelectasis percentage

The following would be the mean atelectasis percentage coverage if a normal distribution were assumed, which is what has been done in some prior studies:

```{r}
data %>%  summarize(
    mean = mean(atelectasis_percent),
    sd = sd(atelectasis_percent)
  ) 
```

And by obesity class:

```{r}
data %>% group_by(type_obesity) %>%
  summarize(
    mean = mean(atelectasis_percent),
    sd = sd(atelectasis_percent)
  ) 
```

As is evident from these numbers, assuming normality causes standard deviation to capture negative values, which is impossible in reality for this variable.

Thus, bootstrapping the mean and 95%CI is expected to lead to more appropriate estimates.

Mean by bootstrapping for the total sample:

```{r}
set.seed(seed)
boot_atel <- one.boot(data$atelectasis_percent, mean, R=10000)
mean_boot <- mean(boot_atel$t)
mean_boot
```

Bootstrap 95% confidence intervals:

```{r}
#| warning: false
boot_ci <- boot.ci(boot_atel)
boot_ci
```

The bias-corrected and accelerated (BCa) bootstrap interval is known to lead to more stable intervals with better coverage. Will report this. However, it is a good thing that here 95%CI through different methods do not lead to widely different results.

Now, I will calculate this for different BMI categories:

```{r}
data_class_1 <- data %>% filter(type_obesity=="Class 1 Obesity") 
data_class_2 <- data %>% filter(type_obesity=="Class 2 Obesity") 
data_class_3 <- data %>% filter(type_obesity=="Class 3 Obesity") 
```

###### Class 1

Mean:

```{r}
set.seed(seed)
boot_class1 <- one.boot(data_class_1$atelectasis_percent, mean, R=10000)
mean_boot_class1 <- mean(boot_class1$t)
mean_boot_class1
```

95% CI:

```{r}
#| warning: false
boot_ci_class1 <- boot.ci(boot_class1)
boot_ci_class1
```

###### Class 2

Mean:

```{r}
set.seed(seed)
boot_class2 <- one.boot(data_class_2$atelectasis_percent, mean, R=10000)
mean_boot_class2 <- mean(boot_class2$t)
mean_boot_class2
```

95% CI:

```{r}
#| warning: false
boot_ci_class2 <- boot.ci(boot_class2)
boot_ci_class2
```

###### Class 3

Mean:

```{r}
set.seed(seed)
boot_class3 <- one.boot(data_class_3$atelectasis_percent, mean, R=10000)
mean_boot_class3 <- mean(boot_class3$t)
mean_boot_class3
```

95% CI:

```{r}
#| warning: false
boot_ci_class3 <- boot.ci(boot_class3)
boot_ci_class3
```

> The mean atelectasis percentage coverage in the sample was `r round(mean_boot,2)`% (95%CI:`r round(boot_ci$bca[1,4],2)`-`r round(boot_ci$bca[1,5],2)`) and according to obesity categories: class 1 (`r round(mean_boot_class1,2)`%, 95%CI:`r round(boot_ci_class1$bca[1,4],2)`-`r round(boot_ci_class1$bca[1,5],2)`), class 2 (`r round(mean_boot_class2,2)`%, 95%CI:`r round(boot_ci_class2$bca[1,4],2)`-`r round(boot_ci_class2$bca[1,5],2)`), and class 3 (`r round(mean_boot_class3,2)`%, 95%CI:`r round(boot_ci_class3$bca[1,4],2)`-`r round(boot_ci_class3$bca[1,5],2)`).

```{r}
#| include: false
rm(boot_atel,mean_boot,boot_ci,data_class_1,data_class_2,data_class_3,boot_class1,mean_boot_class1,boot_ci_class1,boot_class2,mean_boot_class2,boot_ci_class2,boot_class3,mean_boot_class3,boot_ci_class3)
```

##### Atelectasis percentage by obesity class

Now, I will continue assessing atelectasis percentage if assumed to be categorical ordinal:

Mean expected frequency:

```{r}
mean_exp <- data %>% 
  mutate(
    atelectasis_percent=factor(atelectasis_percent)) %>%
  summarize(
    mean_expected_freq = n()/(nlevels(type_obesity)*nlevels(atelectasis_percent))
    )
mean_exp
```

Mean expected frequency is greater than 5.0, so chi-squared without continuity correction is adequate.

Frequencies:

```{r}
frequencies <- table(atelectasis_percent,type_obesity)
frequencies
```

Percentage by obesity class

```{r}
prop_fig2a <- prop.table(frequencies,margin=2)
round(prop_fig2a*100,2)
```

Barplot of absolute frequencies:

```{r}
barplot(frequencies,beside=TRUE)
```

```{r}
#| warning: false
chi <- chisq.test(frequencies, correct=FALSE)
chi
```

##### Barplot of atelectasis percentage by obesity class category

```{r, fig.height=5, fig.width=8}
barplot(prop_fig2a,beside=TRUE,ylim=c(0,1),ylab="Relative frequency",
        col=brewer.pal(9,"Blues"),
        legend.text=c("0%","2.5%","5%","7.5%","10%","12.5%","15%","17.5%","27.5%"),
        space = c(0.2, 1.5)
        )
Figure2a <- recordPlot()
```

```{r}
#| include: false
rm(mean_exp,frequencies,percent,frequencies,chi,Figure2a,prop_fig2a)
```

##### Smooth term?

```{r}
plot(atelectasis_percent~BMI, 
     main="Scatterplot", 
     xlab="Body mass index (kg/m²)", 
     ylab="Atelectasis percent (%)"
     )
```

Atelectasis percent seems to increase as BMI increases. However, relationship is not linear.

```{r}
#| include: false 
## Change 'false' for 'true' above to show plot. 

# Would a smooth term be more useful to model atel_percent? Fit with loess:   
ggplot(data, aes(BMI,atelectasis_percent)) + 
  geom_point(size=0.6,color="gray40") + 
  geom_smooth(method="loess", color="darkblue") +
  ylab("Atelectasis percent (%)") + 
  xlab("Body mass index (kg/m²)")  +
  theme_bw() + 
  theme(panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"),
        axis.text.x = element_text(size=rel(1.2)),
        axis.text.y = element_text(size=rel(1.2))
        )
```

Models evaluated with the accompanying sourced script ***nonlinear_BMI_Atelectasis.R***

```{r}
#| include: false

source("scripts/nonlinear_BMI_Atelectasis.R", local = knitr::knit_global())
```

All models are significantly better than linear. Thus, using a smooth term for BMI to predict atelectasis percent is better than modelling a linear relationship.

Best AIC:

```{r}
data.frame(
  model = c("k=2","k=4","k=6","k=8"),
  AIC = round(c(AIC_k2,AIC_k4,AIC_k6,AIC_k8),1)
) %>% gt()
```

Regarding AIC, greatest improvement in AIC is k=6. Will model with k=5 and k=7 to compare

Best AIC:

```{r}
data.frame(
  model = c("k=5","k=6","k=7"),
  AIC = round(c(AIC_k5,AIC_k6,AIC_k7),1)
) %>% gt()
```

k=6 offers the lowest AIC. Will keep k=6 to model.

```{r}
fig1a
```

Positive non-monotonic relationship since atelectasis increases as BMI increases only after \~BMI equal to 42.

Will assess Spearman's correlation again only to have a rough idea (will not report this in the paper since the relationship is not monotonic):

```{r}
spearman <- cor.test(spo2_VPO,atelectasis_percent,
                     method="spearman",
                     exact=FALSE
                     )
spearman
```

> Atelectasis percent exhibited a negative non-linear non-monotonic relationship with SpO2 (**Figure 1A**, rho= `r round(spearman$estimate,3)`, p\<0.001).

Note that this p-value refers to the smooth term vs linear as assessed in GAM models.

Interestingly, this figure is almost a mirror image of the priorly created plot for SpO2 \~ BMI.

```{r}
#| include: false
rm(AIC_k2,AIC_k3,AIC_k4,AIC_k5,AIC_k6,AIC_k7,AIC_k8,AIC_k10,model,spearman,fig1a)
```

## Atelectasis - age

```{r}
boxplot(age~atelectasis,
        ylab="Age",
        xlab="Atelectasis"
        )
```

Assess distribution of age by atelectasis (yes/no):

```{r}
data_age <- data %>% group_by(atelectasis)
```

```{r}
#| include: false
## Change 'false' for 'true' above to show plot. 

#Assess distribution of age by atelectasis (yes/no):   
ggplot(data_age,aes(x = age)) +
  geom_histogram(fill = "lightsteelblue4", colour = "black") +
  facet_grid(atelectasis ~ .)
```

```{r}
#| include: false 
## Change 'false' for 'true' above to show plot. 

qqPlot(age ~ atelectasis, data=data_age)
```

Distribution near-normal, will assess mean and variance for further testing:

```{r}
mean_age <- data_age %>% 
  summarise(n=n(),
            age_mean = round(mean(age),2),
            sd = round(sd(age),2),
            variance = round(var(age),4)
            )
mean_age %>% gt()
```

Variances near-similar, but group sizes differ. Welch's t-test more suitable:

```{r}
t_test <- t.test(age ~ atelectasis, data = data_age)
t_test
```

> Age was similarly distributed among patients without atelectasis (`r round(mean_age$age_mean[2],1)`, sd:`r round(mean_age$sd[2],1)`) and those with atelectasis (`r round(mean_age$age_mean[1],1)`, sd:`r round(mean_age$sd[1],1)`) (p=`r round(t_test$p.value,3)`).

```{r}
#| include: false 
rm(data_age,mean_age,t_test)
```

## Atelectasis - sex

Mean expected frequency:

```{r}
mean_exp <- data %>% 
  drop_na(sex, atelectasis) %>% 
  summarize(mean_expected_freq = n()/(nlevels(sex)*nlevels(atelectasis)))

mean_exp
```

Mean expected frequency is greater than 5.0, so chi-squared without continuity correction is adequate.

Frequencies:

```{r}
frequencies <- table(sex, atelectasis)
frequencies
```

Percentage:

```{r}
percent <- round((prop.table(frequencies, 1)*100),2)
percent
```

Mosaic Plot

```{r, fig.height=3, fig.width=5}
data %>% 
  mutate(atelectasis = fct_relevel(atelectasis, "No", "Yes")) %>%
ggplot() +
  geom_mosaic(
    aes(x = product(atelectasis,sex), 
        fill=atelectasis),
    na.rm = TRUE
    ) +
  scale_fill_manual(values=c("grey95","lightsteelblue4")) +
  labs(
    y = "Atelectasis",
    x = "Sex"
  ) +
  theme_mosaic() + 
  theme(axis.text.x=element_text(size=rel(0.8)))
```

```{r}
chi <- chisq.test(frequencies, correct=FALSE)
chi
```

> There were no significant differences in atelectasis ocurrence between men (`r round(percent[2,1],1)`%) and women (`r round(percent[1,1],1)`%) (p=`r round(chi$p.value,3)`).

```{r}
#| include: false 
rm(mean_exp,frequencies,percent,chi)
```

## Atelectasis - OSA

Mean expected frequency:

```{r}
mean_exp <- data %>% 
  drop_na(sleep_apnea, atelectasis) %>% 
  summarize(mean_expected_freq = n()/(nlevels(sleep_apnea)*nlevels(atelectasis)))

mean_exp
```

Mean expected frequency is greater than 5.0, so chi-squared without continuity correction is adequate.

Frequencies:

```{r}
frequencies <- table(sleep_apnea, atelectasis)
frequencies
```

Percentage:

```{r}
percent <- round((prop.table(frequencies, 1)*100),2)
percent
```

Mosaic Plot

```{r, fig.height=3, fig.width=5}
data %>% 
  mutate(atelectasis = fct_relevel(atelectasis, "No", "Yes")) %>%
ggplot() +
  geom_mosaic(
    aes(x = product(atelectasis,sleep_apnea), 
        fill=atelectasis),
    na.rm = TRUE
    ) +
  scale_fill_manual(values=c("grey95","lightsteelblue4")) +
  labs(
    y = "Atelectasis",
    x = "Obstructive sleep apnea"
  ) +
  theme_mosaic() + 
  theme(axis.text.x=element_text(size=rel(0.8)))
```

```{r}
chi <- chisq.test(frequencies, correct=FALSE)
chi
```

> Patients with a diagnosis of obstructive sleep apnea had atelectasis more frequently (`r round(percent[2,1],1)`%) than those without the diagnosis (`r round(percent[1,1],1)`%) (p\<0.001).

```{r}
#| include: false 
rm(mean_exp,frequencies, percent, chi)
```

#### Atelectasis location by OSA

Mean expected frequency:

```{r}
mean_exp <- data %>% 
  drop_na(sleep_apnea, atelectasis_location) %>% 
  summarize(mean_expected_freq = n()/(nlevels(sleep_apnea)*nlevels(atelectasis_location)))

mean_exp
```

Mean expected frequency is greater than 5.0, so chi-squared without continuity correction is adequate.

Frequencies:

```{r}
frequencies <- table(sleep_apnea, atelectasis_location)
frequencies
```

Percentage:

```{r}
percent <- round((prop.table(frequencies, 1)*100),2)
percent
```

Mosaic Plot

```{r, fig.height=3, fig.width=5}
ggplot(data = data) +
  geom_mosaic(aes(
    x = product(atelectasis_location,sleep_apnea),
    fill=atelectasis_location),
    na.rm = TRUE
    ) +
  scale_fill_manual(values=c("seagreen1","seagreen4")) +
  labs(
    y = "Atelectasis location",
    x = "Obstructive sleep apnea"
  ) +
  theme_mosaic() 
```

```{r}
chi <- chisq.test(frequencies, correct=FALSE)
chi
```

> The location of atelectasis was not different among patients with and without OSA (p=`r round(chi$p.value,3)`).

```{r}
#| include: false 
rm(mean_exp,frequencies, percent, chi)
```

## Atelectasis - SpO2

```{r}
data_spo2 <- data %>% group_by(atelectasis) 

median_spo2 <- data_spo2 %>% summarize(n = n(),
                                       spo2_median = median(spo2_VPO), 
                                       Q1 = quantile(spo2_VPO,0.25), 
                                       Q3 = quantile(spo2_VPO,0.75), 
                                       min = min(spo2_VPO), 
                                       max = max(spo2_VPO)
                                       )
median_spo2 %>% gt()
```

```{r}
boxplot(spo2_VPO ~ atelectasis,
        ylab="SpO2 (%)",
        xlab="Atelectasis"
        )
```

Distribution not normal and influential outliers. Will assess non-parametrically.

```{r}
wil <- wilcox.test(spo2_VPO ~ atelectasis, 
                   data = data_spo2, 
                   exact = FALSE
                   )
wil
```

> The median SpO2 was significantly lower in patients with atelectasis (`r round(median_spo2$spo2_median[1],1)`, IQR: `r round(median_spo2$Q1[1],1)`-`r round(median_spo2$Q3[1],1)`) compared to those without (`r round(median_spo2$spo2_median[2],1)`, IQR: `r round(median_spo2$Q1[2],1)`-`r round(median_spo2$Q3[2],1)`) (p\<0.001).

```{r}
#| include: false 
rm(data_spo2,median_spo2,wil)
```

## Atelectasis location - SpO2

```{r}
data_spo2 <- data %>% 
  group_by(atelectasis_location) %>% 
  drop_na(atelectasis_location)

median_spo2 <- data_spo2 %>% summarize(n = n(),
                                       spo2_median = median(spo2_VPO), 
                                       Q1 = quantile(spo2_VPO,0.25), 
                                       Q3 = quantile(spo2_VPO,0.75), 
                                       min = min(spo2_VPO), 
                                       max = max(spo2_VPO)
                                       )
median_spo2 %>% gt()
```

```{r}
boxplot(spo2_VPO ~ atelectasis_location,
        ylab="SpO2 (%)",
        xlab="Atelectasis location"
        )
```

Distribution not normal and likely influential outliers. Will assess non-parametrically.

```{r}
wil <- wilcox.test(spo2_VPO ~ atelectasis_location, 
                   data = data_spo2, 
                   exact = FALSE
                   )
wil
```

> The median SpO2 was significantly lower in patients with bilateral atelectasis (`r round(median_spo2$spo2_median[2],1)`, IQR: `r round(median_spo2$Q1[2],1)`-`r round(median_spo2$Q3[2],1)`) compared to those with unilateral atelectasis (`r round(median_spo2$spo2_median[1],1)`, IQR: `r round(median_spo2$Q1[1],1)`-`r round(median_spo2$Q3[1],1)`) (p=`r round(wil$p.value,3)`).

```{r}
#| include: false 
rm(data_spo2,median_spo2,wil)
```

## Atelectasis percent - SpO2

### Smooth term?

Scatterplot

```{r}
plot(spo2_VPO~atelectasis_percent, data=data, 
     main="Scatterplot", 
     xlab="Atelectasis percent (%)", 
     ylab="SpO2 (%)")
```

Decreasing SpO2 as atelectasis percent increases.

Would a smooth term be more useful to model SpO2?

```{r}
#| include: false 
## Change 'false' for 'true' above to show plot. 

# Would a smooth term be more useful to model SpO2? Fit with loess:   

ggplot(data, aes(atelectasis_percent,spo2_VPO)) + 
  geom_point(size=0.6,color="gray40") + 
  geom_smooth(method="loess", color="black") +
  ylab("SpO2 (%)") + xlab("Atelectasis percent (%)") + 
  ylim(85,100) +
  theme_bw() + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(colour = "black"),
        axis.text.x = element_text(size=rel(1.2)), 
        axis.text.y = element_text(size=rel(1.2))
        )
```

Models evaluated with the accompanying sourced script ***nonlinear_Atelectasis_SpO2.R***

```{r}
#| include: false

source("scripts/nonlinear_Atelectasis_SpO2.R", local = knitr::knit_global())
```

All models are significantly better than linear. Thus, using a smooth term for atelectasis percent is better than modelling a linear relationship.

Best AIC:

```{r}
data.frame(
  model = c("k=2","k=4","k=6","k=8"),
  AIC = round(c(AIC_k2,AIC_k4,AIC_k6,AIC_k8),1)
) %>% gt()
```

Regarding AIC, no model offers greater improvement in AIC than k=6. Will try a model with k=5.

Best AIC:

```{r}
data.frame(
  model = c("k=4","k=5","k=6"),
  AIC = round(c(AIC_k4,AIC_k5,AIC_k6),1)
) %>% gt()
```

There is a drop in AIC for k=5, which also offers the best k-index. Nonetheless, one problem with this is that the extra knot is explaining a clump around 12.%, for which there was only one single observation. Thus, it is likely that this clump and additional knot is only explaining noise in the data, and would thus not be a good representation of the trend in the variable. Thus, will keep k=4 to model as this model offers the best visual representation of the trend in all categories.

```{r}
fig1c
```

Negative monotonic relationship since SpO2 decreases as BMI increases. Will assess Spearman's correlation coefficient to report in paper:

```{r}
spearman <- cor.test(spo2_VPO,atelectasis_percent,
                     method="spearman",
                     exact=FALSE
                     )
spearman
```

> Atelectasis percent exhibited a negative non-linear monotonic relationship with SpO2 (**Figure 1C**, rho= `r round(spearman$estimate,3)`, p\<0.001).

### Figure 1

Created with the accompanying sourced script ***Figure1.R***

```{r}
#| include: false

source("scripts/Figure1.R", local = knitr::knit_global())
```

```{r, fig.height=8, fig.width=10}
grid.arrange(fig1a, blank_plot, fig1b, fig1c, ncol=2)
```

```{r}
#| include: false
rm(AIC_k2,AIC_k4,AIC_k5,AIC_k6,AIC_k8,model,spearman,figure1,fig1a,fig1b,fig1c,blank_plot)
```

```{r}
#| include: false
detach(data)
```

### Ordinal variable

Since there is only one participant in the 30% category, will collapse with the 20 category for further analyses:

```{r}
datamodel <- data
datamodel$atelectasis_percent <- factor(
  datamodel$atelectasis_percent,
  levels=c(0,2.5,5,7.5,10,12.5,15,17.5,27.5)
  ) %>% 
  fct_collapse("17.5%" = c(17.5,27.5)) %>% 
  factor(labels = c("0%","2.5%","5%","7.5%","10%","12.5%","15%","17.5%"))

table(datamodel$atelectasis_percent)
```

```{r}
#| include: false 
data_spo2 <- datamodel %>% group_by(atelectasis_percent)
```

```{r}
#| include: false 
## Change 'false' for 'true' above to show plot. 

# Assess distribution of SpO2 by atelectasis percent categories: 
ggplot(data_spo2,aes(x = spo2_VPO)) +
  geom_histogram(fill = "aquamarine", colour = "black") +
  facet_grid(atelectasis_percent ~ .)
```

Distribution not normal, group sizes are different and there are outliers in both directions, depending where you are located. Thus, will proceed with non-parametric assessment:

```{r}
median_spo2 <- data_spo2 %>% 
  summarize(n = n(), 
            spo2_median = median(spo2_VPO), 
            Q1 = quantile(spo2_VPO,0.25), 
            Q3 = quantile(spo2_VPO,0.75), 
            min = min(spo2_VPO), 
            max = max(spo2_VPO)
            )
median_spo2 %>% gt()
```

```{r}
krus <- kruskal.test(spo2_VPO ~ atelectasis_percent, data = data_spo2)
krus
```

> There was a decreasing trend in median SpO2 with higher atelectasis percentage extension (p\<0.001).

```{r}
ggplot(datamodel, aes(x = atelectasis_percent, y = spo2_VPO)) + 
  geom_boxplot(width = .4, outlier.shape = NA, fill="aliceblue") +
  geom_point(size = 1.5, alpha = .3, 
             position = position_jitter(seed = 1, width = .1)) + 
  ylab("SpO2 (%)") + xlab("Atelectasis percent") + 
  ylim(85,100) +
  labs(tag="Kruskall-Wallis: p<0.001") + 
  theme_bw() + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(colour = "black"),
        axis.ticks.length.x = unit(.1, "in"),
        axis.text.x = element_text(size=rel(1.2)),
        axis.text.y = element_text(size=rel(1.2)),
        plot.tag.position = "top", 
        plot.tag = element_text(size=rel(0.9))
        )
```

```{r}
#| include: false 
rm(data_spo2,median_spo2,krus) 
```

## Relationship between OSA, obesity type and atelectasis percent:

```{r}
#| include: false 
## Change 'false' for 'true' above to show table 

# This also gives an idea of which sizes to use for next plot: 
datamodel %>% 
  group_by(atelectasis_percent, sleep_apnea, type_obesity) %>% 
  summarize(
    n = n()
  )
```

```{r}
datamodel %>% mutate(
  type_obesity = fct_recode(
    type_obesity,
    "1" = "Class 1 Obesity",
    "2" = "Class 2 Obesity",
    "3" = "Class 3 Obesity"
    )
  ) %>% 
  count(type_obesity, sleep_apnea, atelectasis_percent) %>%
  ggplot(
    aes(x = sleep_apnea,
        y = atelectasis_percent,
        color = sleep_apnea)
    ) +
  geom_point(aes(group = sleep_apnea, size = n)) +
  scale_color_manual(values=c("gray40","darkred")) +
  facet_wrap(~type_obesity, scales = "free_x",
             labeller = labeller(type_obesity = label_both)
             ) +
  scale_size(breaks = c(1, 5, 10, 20, 30, 40, 50, 60)) + 
  labs(
    y = "Atelectasis percent",
    x = "Obstructive sleep apnea"
  ) +
  theme_light()
```

Sleep apnea was more common with higher BMI categories and also with higher atelectasis percentage. Atelectasis percent increases at higher obesity classes.

```{r}
#| include: false

# Run this chunk if you wish to clear your environment and unload packages.

rm(data, datamodel, figfolder, seed)

pacman::p_unload(negate = TRUE)
```
