---
title: "Preoperative Atelectasis"
subtitle: "Part 1: Overview, selection criteria, and missing data"
author: "Javier Mancilla Galindo"
date: "`r Sys.Date()`"
execute: 
  echo: true
  warning: false
format:
  html:
    toc: true
    toc_float: true
    embed-resources: true
editor: visual
---

# Setup

#### Packages used

```{r}
if (!require("pacman", quietly = TRUE)) {
  install.packages("pacman")
}

pacman::p_load(
  tidyverse, # Used for basic data handling and visualization.
  dataverse, # Retrieve dataset from the Harvard dataverse. 
  overviewR, # Used to assess missing data.
  table1 #Used to add labels to variables.
)
```

```{r}
#| include: false  

# Create directories for sub-folders  
psfolder <- "../data/processed"
reports <- "../docs/reports"
dir.create(psfolder, showWarnings = FALSE)
dir.create(reports, showWarnings = FALSE)
```

##### Session info

```{r}
#| echo: false  

# Credits chunk of code: Alex Bossers, Utrecht University (a.bossers@uu.nl)

# remove clutter
session <- sessionInfo()
session$BLAS <- NULL
session$LAPACK <- NULL
session$loadedOnly <- NULL
# write log file
writeLines(
  capture.output(print(session, locale = FALSE)),
  paste0("sessions/",lubridate::today(), "_session_Part_1.txt")
)

session

rm(session)
```

```{r}
#| include: false  

# Load dataset  
data <- get_dataframe_by_name(
  filename = "atelectasis_prevalence.tab",
  dataset = "10.7910/DVN/4JZZLB", 
  server = "dataverse.harvard.edu")


# Recode variables  
source("scripts/variable_names_raw.R", local = knitr::knit_global())
```


```{r}
#| include: false  

# Can be used to check the structure of the dataset after prior specification.
str(data)
```

# General overview

```{r}
summary(data)
```

### Exclude participants with CO-RADS ≥ 3:

Participants with higher probability of having a current diagnosis of COVID-19 are expected to have chest CT alterations due to COVID-19 pneumonia. Thus, will be excluded.

Number of patients according to CO-RADS:

```{r}
summary(data$CORADS)
```

```{r}
count(data)
```

```{r}
data <- data %>%
  filter(as.numeric(CORADS) < 3) %>%
  droplevels(data$CORADS)
```

```{r}
count(data)
```

### Exclude participants with prior COVID-19:

Since prior COVID-19 is considered a confounder and since there are only 3 participants with prior COVID-19 which would provide difficult to assess the role of prior COVID-19 in analyses, participants with prior COVID-19 were excluded from the analysis.

```{r}
count(data)
```

```{r}
data <- data %>%
  filter(prior_covid19 == "No") %>%
  droplevels(data$prior_covid19)
```

```{r}
count(data)
```

Will remove prior_covid19 column as it no longer provides information of a varying characteristic. Similarly, rapid_covid_test does not provide additional information.

```{r}
length(data)
```

```{r}
data <- data %>% select(-c(prior_covid19, rapid_covid_test))
```

```{r}
length(data)
```

# Missing data per variable

```{r}
overview_na(data)
```

Missing data for PCR_covid and is explained since only patients who decided to have a test performed on their own will reported the result. The medical center did not require a negative PCR test at that time during the pandemic, reason why PCR tests were not systematically performed. As shown in earlier summary of variables, all available tests (n=3) were negative. This variable will not be analyzed further downstream:

```{r}
data <- data %>% select(-PCR_covid)
```

The variable atelectasis_location has missing data since those patients who did not have atelectasis naturally do not have a location recorded. Will assess if data are missing for those who had atelectasis:

```{r}
data %>%
  filter(atelectasis == "Yes") %>%
  group_by(atelectasis_location) %>%
  overview_na()
```

There is no missing data for ***atelectasis_location*** after sub-setting only those who had atelectasis.

Lastly, I will subset all participants without the prior variable to further assess the extent of missing data for other variables:

```{r}
data %>%
  select(-c(atelectasis_location)) %>%
  overview_na()
```

Missing data constitutes **less than 5%** for all remaining variables. Thus, will proceed with **complete-case analysis** without performing any data imputation procedure.

```{r}
#| include: false  

#Save dataset
write.csv(data, 
          file = paste0(psfolder,"/atelectasis_included.csv"),
          row.names=FALSE)
```

```{r}
#| include: false  

# This chunk removes all created objects and loaded packages to keep
# your environment and session clean after running this document. 

rm(data, psfolder, reports, variable_names_raw)
pacman::p_unload(negate = TRUE)

```
