---
title: "Preoperative Atelectasis"
subtitle: "Part 3: Assessment of Independent Variables"
author: "Javier Mancilla Galindo"
date: "`r Sys.Date()`"
project: 
    output-dir: ../docs/reports
execute: 
  echo: true
  warning: false
format:
  html:
    toc: true
    toc_float: true
    embed-resources: true
  pdf: 
    prefer-html: true
editor: visual
---

# Setup 

#### Packages used   
```{r}
if (!require("pacman", quietly = TRUE)) {
  install.packages("pacman")
}


pacman::p_load(
  tidyverse, # Used for basic data handling and visualization.
  table1, #Used to add lables to variables.
  RColorBrewer, #Color palettes for data visualization. 
  gridExtra, #Used to arrange multiple ggplots in a grid.  
  grid, #Used to arrange multiple ggplots in a grid.
  flextable, #Used to export tables.  
  officer,  #Used to export tables.
  mgcv, #Used to model non-linear relationships with a general additive model.  
  ggmosaic, #Used to create mosaic plots.   
  car, #Used to visualize distribution of continuous variables (stacked Q-Q plots).
  dagitty #Used in conjunction with https://www.dagitty.net/ to create directed acyclic graph to inform statistical modelling.
)
```


##### Session   
```{r}
#| echo: false

# Credits chunk of code: Alex Bossers, Utrecht University (a.bossers@uu.nl)

# remove clutter
session <- sessionInfo()
session$BLAS <- NULL
session$LAPACK <- NULL
session$loadedOnly <- NULL
# write log file
writeLines(
  capture.output(print(session, locale = FALSE)),
  paste0("sessions/",lubridate::today(), "_session_Part_3.txt")
)

session

rm(session)

```


```{r}
#| include: false

# Create directories for sub-folders 
figfolder <- "../results/output_figures"
tabfolder <- "../results/output_tables"
dir.create(figfolder, showWarnings = FALSE)
dir.create(tabfolder, showWarnings = FALSE)
```


```{r}
#| include: false

# Load dataset  
data <- read.csv("../data/processed/atelectasis_included.csv",
                 na.strings="NA", 
                 row.names = NULL)
# Recode variables 
source("scripts/variable_names.R")

```

# Assessment of independent variables     

The selection of variables that will be assessed is according to the following directed acyclic graph which will be used again before statistical modelling, to assess conditional independencies.     

## DAG

DAG generated in the [DAGitty website](https://www.dagitty.net/) and sourced from the accompanying script ***DAG.R***   
```{r}
#| echo: false

source("scripts/DAG.R")
plot(DAG)
```

Other variables that are potential confounders are not shown in this DAG since they were addressed by design in this study as follows:   
- Current COVID-19: Exclusion criteria were applied to **n=2** patients with CO-RADS 3 and **n=2** with CO-RADS 4. Only participants with low probability of COVID-19 (CO-RADS 1 and 2) were included in this study.      
- Prior COVID-19: This was an exclusion criterion (**n=3**).   
- Bronchiectasis: This was an exclusion criterion (n=0).      

## Description of independent variables  

#### Age  
```{r}
#| include: false

attach(data)
```

```{r}
summary(age)
```

> The mean age was `r round(mean(data$age, na.rm=TRUE),1)` (SD: `r round(sd(data$age, na.rm=TRUE),2)`).     

#### Sex  
```{r}
frequencies <- table(sex)
frequencies
```

```{r}
percentage <- round(prop.table(frequencies)*100,1)
percentage
```

> Most patients in the sample were woman (n=`r frequencies[1]`, `r percentage[1]`%).

```{r, include=FALSE}
#| include: false
rm(frequencies,percentage)
```

#### Body mass index (BMI)  
```{r}
summary(BMI)
```


```{r}
frequencies <- table(type_obesity)
frequencies
percentage <- round(prop.table(frequencies)*100,1)
percentage

```
Distribution of BMI was assessed earlier. It is right-skewed due to extreme values (verified outliers). The WHO classification of BMI for obesity class will be used to complement descriptions and for potential use later during statistical modelling.     

> The median BMI was `r median(data$BMI)` (IQR: `r quantile(data$BMI, 0.25)`- `r quantile(data$BMI, 0.75)`). The distribution of BMI was right-skewed due to extreme BMI values (range: `r min(data$BMI)`- `r max(data$BMI)`). Most patients were in the class 3 obesity category (n=`r tab[3]`, `r prop[3]`%), followed by class 1 (n=`r tab[1]`, `r prop[1]`%) and 2 (n=`r tab[2]`, `r prop[2]`%). 

```{r, include=FALSE}
rm(tab,prop)
```

#### SpO2  
```{r}
summary(spo2_VPO)
```
Distribution of SpO2 during the pre-anesthetic assessment was presented earlier. It is left-skewed due to some participants exhibiting decreased SpO2. Will categorize according to clinical categories to assess the proportion of patients with decreased SpO2.   

###### Proportion of patients with decreased SpO2

First, create SpO2 breaks:   
```{r}
data <- data %>% 
  mutate(spo2_cat = cut(spo2_VPO,
                        breaks=c(87,90,94,100),
                        right=TRUE,
                        labels=c("≤90","90 to 94",">94")
                        )
         )
```


```{r, include=FALSE}
detach(data)
attach(data)
# This is done to update the attached dataset with the newly created variable.   
```


```{r}
tab <- table(spo2_cat)
tab
prop <- round(prop.table(tab)*100,1)
prop
```

> The median SpO2 during the pre-anethetic assessment was `r median(data$spo2_VPO)` (IQR: `r quantile(data$spo2_VPO, 0.25)`-`r quantile(data$spo2_VPO, 0.75)`) %, with a minimum value of `r min(data$spo2_VPO)`%. Of these,  n=`r tab[3]` (`r prop[3]`%) had normal SpO2 (above 94%), whereas n=`r tab[2]` (`r prop[2]`%) had a value in the 90-94% range, and n=`r tab[1]` (`r prop[1]`%) had ≤90%.

```{r, include=FALSE}
rm(tab,prop)
```

#### Obstructive sleep apnea  
```{r}
tab <- table(sleep_apnea)
tab
prop <- round(prop.table(tab)*100,1)
prop
```
> Patients with a diagnosis of OSA were `r prop[2]`% (n=`r tab[2]`) of the sample.  

```{r, include=FALSE}
rm(tab,prop)
```

#### Altitude   
```{r}
summary(altitude)
```
Distribution of altitude was assessed earlier. Cannot assume normal distribution. Thus, I will create a new variable categorizing values according to the [study by Crocker ME, et al](https://doi.org/10.1016/S2214-109X(19)30543-1).   

```{r}
data <- data %>% 
  mutate(altitude_cat = cut(altitude,
                            breaks=c(0,1000,2500),
                            right=FALSE,
                            labels=c("Low altitude","Moderate altitude")
                            )
         )
```

```{r, include=FALSE}
detach(data)
attach(data)
# This is done to update the attached dataset with the newly created variable.   
```

```{r}
tab <- table(altitude_cat)
tab
round(prop.table(tab)*100,1)
```

```{r, include=FALSE}
rm(tab)
```


#### Hemoglobin     
```{r}
summary(hb)
```
Distribution of hemoglobin was assessed earlier. Two participants don't have a hemoglobin value.     

## Relationships between independent variables  

> Characteristics of participants according to BMI class are shown in **Table 1**.  

### BMI and SpO2   
Scatterplot  
```{r}
plot(spo2_VPO~BMI, 
     main="Scatterplot", 
     xlab="Body mass index (kg/m²)", 
     ylab="SpO2 (%)"
     )
```

Relationship does not seem to be linear (also, variables were not normally distributed, with outliers), but suggests a negative correlation. Will assess if a smooth BMI term explains SpO2 better, and if so, what is the best number of knots to model this relationship:       

GAM model with k=2:   
```{r}
model<-gam(spo2_VPO~s(BMI,k=2))
AIC_k2 <- AIC(model)
summary(model)
k.check(model)
plot(model)
```

GAM model with k=4:   
```{r}
model<-gam(spo2_VPO~s(BMI,k=4))
AIC_k4 <- AIC(model)
summary(model)
k.check(model)
plot(model)
```


GAM model with k=6:   
```{r}
model<-gam(spo2_VPO~s(BMI,k=6))
AIC_k6 <- AIC(model)
summary(model)
k.check(model)
plot(model)
```

GAM model with k=8:   
```{r}
model<-gam(spo2_VPO~s(BMI,k=8))
AIC_k8 <- AIC(model)
summary(model)
k.check(model)
plot(model)
```

GAM model with k=12:   
```{r}
model<-gam(spo2_VPO~s(BMI,k=12))
AIC_k12 <- AIC(model)
summary(model)
k.check(model)
plot(model)
```


Best AIC:  
```{r}
list(AIC_k2,AIC_k4,AIC_k6,AIC_k8,AIC_k12)
```
All models are significantly better than linear. Thus, using a smooth term for BMI is better than modelling a linear relationship.   

Regarding AIC, the models with k>6 are not better at explaining the variance. Will try a model with k=5 since the best model is expected to be anywhere between k=4 and k=6:   

GAM model with k=5 (4 degrees of freedom):   
```{r}
model<-gam(spo2_VPO~s(BMI,k=5))
AIC_k5 <- AIC(model)
summary(model)
k.check(model)
plot(model)
```

Best AIC:  
```{r}
list(AIC_k4,AIC_k5,AIC_k6)
```

Model with k=5 still offers and advantage compared to k=4 (drop in AIC). No other improvements in k-index or visual representation are achieved with higher k. Thus, will use k=5 to model.   

```{r}
fig1b <- ggplot(data, aes(BMI,spo2_VPO)) + 
  geom_point(size=0.6,color="gray60") +
  geom_smooth(method="gam",formula = y ~ s(x, bs = "cs", k = 5), color="cadetblue4")  + 
  labs(y="SpO2 (%)",x="Body mass index (kg/m²)",tag="B") + 
  ylim(83,100) + xlim(30,80) +
  theme_bw() + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"),
        axis.text.x = element_text(size=rel(1.2)), 
        axis.text.y = element_text(size=rel(1.2))
        )
fig1b
```
   
Negative non-monotonic relationship since SpO2 decreases, but then seems to increase slightly again at BMI 40, followed by a marked decrease as BMI decreases at values higher than ~42. 

Spearman's correlation coefficient shouldn't be used due to relationship not being monotonically decreasing. However, I will calculate it just to have a rough idea (but will not report this in the paper).        
```{r}
spearman <- cor.test(spo2_VPO,BMI,method="spearman",exact=FALSE)
spearman
```

##### Figure 1b 
```{r}
ggsave("Figure1b.png",plot=fig1b,
       path=figfolder, 
       width = 8,  height = 6, units = "in", dpi = 300
       )
```


> BMI exhibited a negative non-linear monotonic relationship with SpO2 (**Figure 1B**, rho= `r round(spearman$estimate,3)`, p<0.001).

```{r, include=FALSE}
rm(AIC_k2,AIC_k4,AIC_k5,AIC_k6,AIC_k8,AIC_k12,model,spearman)
```

### BMI and age  
Scatterplot  
```{r}
plot(age~BMI, 
     main="Scatterplot", 
     ylab="Age (years)", 
     xlab="Body mass index (BMI)"
     )
```

Datapoints scattered. Relationship probably linear, but there are influential true outliers with extreme BMI. Will assess with Spearman correlation analysis due to extreme BMI values.  

```{r}
spearman <- cor.test(age,BMI,method="spearman",exact=FALSE)
spearman
```

> Age had a weak negative correlation with BMI (rho= `r round(spearman$estimate,3)`, p=`r round(spearman$p.value,3)`).  

```{r, include=FALSE}
rm(spearman)
```


### BMI and sex   

```{r}
data_BMI <- data %>% group_by(sex) 

median_bmi <- data_BMI %>% 
  summarize(n = n(), 
            BMI_median = median(BMI), 
            Q1 = quantile(BMI,0.25), 
            Q3 = quantile(BMI,0.75), 
            min = min(BMI), 
            max = max(BMI)
            )
median_bmi
```

```{r}
boxplot(BMI ~ sex)
```

Distribution not normal and influential outliers. Will assess non-parametrically.  

```{r}
wil <- wilcox.test(BMI ~ sex, data = data_BMI, exact = FALSE)
wil
```

> The median BMI was not different between men (`r round(median_bmi$BMI_median[2],1)`, IQR: `r round(median_bmi$Q1[2],1)`-`r round(median_bmi$Q3[2],1)`) and women (`r round(median_bmi$BMI_median[1],1)`, IQR: `r round(median_bmi$Q1[1],1)`-`r round(median_bmi$Q3[1],1)`) (p=`r round(wil$p.value,3)`).  

```{r, include=FALSE}
rm(data_BMI,median_bmi,wil)
```


### BMI and sleep apnea  
```{r}
boxplot(BMI ~ sleep_apnea)
```

Distribution not normal and influential outliers. Will assess non-parametrically.  

```{r}
data_BMI <- data %>% group_by(sleep_apnea) 

median_bmi <- data_BMI %>% 
  summarize(n = n(),
            BMI_median = median(BMI),
            Q1 = quantile(BMI,0.25),
            Q3 = quantile(BMI,0.75),
            min = min(BMI),
            max = max(BMI)
            )
median_bmi
```

```{r}
wil <- wilcox.test(BMI ~ sleep_apnea, data = data_BMI, exact = FALSE)
wil
```

> The median BMI was significantly higher in participants with sleep apnea (`r round(median_bmi$BMI_median[2],1)`, IQR: `r round(median_bmi$Q1[2],1)`-`r round(median_bmi$Q3[2],1)`) compared to those without OSA (`r round(median_bmi$BMI_median[1],1)`, IQR: `r round(median_bmi$Q1[1],1)`-`r round(median_bmi$Q3[1],1)`) (p=`r round(wil$p.value,3)`).  

```{r, include=FALSE}
rm(data_BMI,median_bmi,wil)
```

### Age and SpO2   

Scatterplot
```{r}
plot(spo2_VPO~age, 
     main="Scatterplot",
     xlab="Age",
     ylab="SpO2 (%)"
     )
```

Do not seem to be correlated. Will apply Spearman's correlation test:  
```{r}
spearman <- cor.test(spo2_VPO,age,method="spearman",exact=FALSE)
spearman
```

> Age and SpO2 were not correlated (rho= `r round(spearman$estimate,3)`, p=`r round(spearman$p.value,3)`).

```{r, include=FALSE}
rm(spearman)
```

### Age and sex  

```{r}
boxplot(age~sex)
```

Assess distribution of age by sex:    

```{r}
data_age <- data %>% group_by(sex)

ggplot(data_age,aes(x = age)) +
  geom_histogram(fill = "firebrick3", colour = "black") +
  facet_grid(sex ~ .)

qqPlot(age ~ sex, data=data_age)
```

Distribution near-normal, but light tails for women. However, t-test could be robust to deviations from normality and differences in group size. Will assess mean and variance for further testing:  
```{r}
mean_age <- data_age %>% 
  summarise(n=n(),
            age_mean = mean(age),
            sd = sd(age),
            variance = var(age)
            )
mean_age
```  

Variances are similar. However, group sizes differ my 10x. Welch's t-test more suitable:    

```{r}
t_test <- t.test(age ~ sex, data = data_age)
t_test
```

> Mean age was similar bethween men (`r round(mean_age$age_mean[2],1)`, sd:`r round(mean_age$sd[2],1)`) and women (`r round(mean_age$age_mean[1],1)`, sd:`r round(mean_age$sd[1],1)`) (p=`r round(t_test$p.value,3)`). 

```{r, include=FALSE}
rm(data_age,mean_age,t_test)
```


### Age and sleep apnea  

```{r}
data_age <- data %>% group_by(sleep_apnea)

ggplot(data_age, aes(x = age, fill=sleep_apnea)) +
  geom_histogram(position = "identity", alpha = 0.4)

qqPlot(age ~ sleep_apnea)
```

Distribution near-normal. Will assess mean and variance for further testing.  
```{r}
mean_age <- data_age %>% 
  summarise(n=n(), 
            age_mean = mean(age), 
            sd = sd(age), 
            variance = var(age)
            )
mean_age
```  

Size per group very different, variances do not look similar. Welch's t-test more suitable:    


```{r}
t_test <- t.test(age ~ sleep_apnea, data = data_age)
t_test
```

> Age was not significantly different between participants with OSA (`r round(mean_age$age_mean[2],1)`, sd:`r round(mean_age$sd[2],1)`) and those without (`r round(mean_age$age_mean[1],1)`, sd:`r round(mean_age$sd[1],1)`) (p=`r round(t_test$p.value,3)`). 

```{r, include=FALSE}
rm(data_age,mean_age,t_test)
```


### SpO2 and sex  

```{r}
boxplot(spo2_VPO ~ sex)
```

```{r}
data_spo2 <- data %>% group_by(sex)

ggplot(data_spo2, aes(x = spo2_VPO, fill=sex)) +
  geom_histogram(position = "identity", alpha = 0.4)

qqPlot(spo2_VPO ~ sex)
```

Distribution deviates from normal and small group size for men. Will assess non-parametrically.  
```{r}
median_spo2 <- data_spo2 %>% 
  summarize(n = n(),
            spo2_median = median(spo2_VPO), 
            Q1 = quantile(spo2_VPO,0.25), 
            Q3 = quantile(spo2_VPO,0.75), 
            min = min(spo2_VPO), 
            max = max(spo2_VPO)
            )
median_spo2
```  

```{r}
wil <- wilcox.test(spo2_VPO ~ sex, data = data_spo2, exact = FALSE)
wil
```

> The median SpO2 was not different between men (`r round(median_spo2$spo2_median[2],1)`, IQR: `r round(median_spo2$Q1[2],1)`-`r round(median_spo2$Q3[2],1)`) and women (`r round(median_spo2$spo2_median[1],1)`, IQR: `r round(median_spo2$Q1[1],1)`-`r round(median_spo2$Q3[1],1)`) (p=`r round(wil$p.value,3)`).  

```{r, include=FALSE}
rm(data_spo2,median_spo2,wil)
```

### SpO2 and sleep apnea  

```{r}
data_spo2 <- data %>% group_by(sleep_apnea)

ggplot(data_spo2, aes(x = spo2_VPO, fill=sleep_apnea)) +
  geom_histogram(position = "identity", alpha = 0.4)

qqPlot(spo2_VPO ~ sleep_apnea)
```

```{r}
boxplot(spo2_VPO ~ sleep_apnea)
```

Distribution not normal, and smaller group size for those with sleep apnea. Will assess non-parametrically.   
```{r}
median_spo2 <- data_spo2 %>% 
  summarize(n = n(), 
            spo2_median = median(spo2_VPO), 
            Q1 = quantile(spo2_VPO,0.25), 
            Q3 = quantile(spo2_VPO,0.75), 
            min = min(spo2_VPO), 
            max = max(spo2_VPO)
            )
median_spo2
```  

```{r}
wil <- wilcox.test(spo2_VPO ~ sleep_apnea, data = data_spo2, exact = FALSE)
wil
```

> Patients with sleep apnea had a lower median SpO2 (`r round(median_spo2$spo2_median[2],1)`, IQR: `r round(median_spo2$Q1[2],1)`-`r round(median_spo2$Q3[2],1)`) than those without OSA (`r round(median_spo2$spo2_median[1],1)`, IQR: `r round(median_spo2$Q1[1],1)`-`r round(median_spo2$Q3[1],1)`) (p<0.001).  

```{r, include=FALSE}
rm(data_spo2,median_spo2,wil)
```


### SpO2 and altitude    
Scatterplot  
```{r}
plot(spo2_VPO~altitude, data=data, 
     main="Scatterplot", 
     xlab="Mean altitude (meters)", 
     ylab="SpO2 (%)"
     )
abline(lm(spo2_VPO~altitude),col="steelblue")
```

There does not seem to be a pattern.          

Would a smooth term be useful to model altitude?   
```{r}
ggplot(data, aes(altitude,spo2_VPO)) + 
  geom_point(size=0.6,color="gray40") + 
  geom_smooth(method="loess", color="cadetblue4") +
  ylab("SpO2 (%)") + xlab("Mean altitude (meters)") + 
  ylim(85,100) + xlim(0,2000) +
  theme_bw() + 
  theme(panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"),
        axis.text.x = element_text(size=rel(1.2)), 
        axis.text.y = element_text(size=rel(1.2))
        )
```

It is likely that a smooth term for SpO2 would be non-informative since there is no clear reasonable pattern in this smooth plot. Additionally, it is well known that any impacts in SpO2 due to altitudes up to 2000 are very limited (i.e 1 to 2 units). [REF](https://thorax.bmj.com/content/73/8/776).  

I will still check if a smooth term may be better than linear in case that adjustment for this variable is needed.   

GAM model with k=2:   
```{r}
model<-gam(spo2_VPO~s(altitude,k=2))
summary(model)
k.check(model)
plot(model)
```

```{r}
model<-gam(spo2_VPO~s(altitude,k=4))
summary(model)
k.check(model)
plot(model)
```

Smooth term is not significantly better than one assuming linearity. Furthermore, the relationship with SpO2 in smooth term does not make any sense (i.e., according to prior reference, SpO2 should decrease at higher altitudes). Thus, it would be very likely that including this term would only explain noise in any case, not the true known causal relationship between SpO2 and altitude.   

Lastly, will check the pattern according to altitude categories, which may be a better term to use in models in any case.     

```{r}
boxplot(spo2_VPO ~ altitude_cat)
```

```{r}
data_spo2 <- data %>% group_by(altitude_cat)

ggplot(data_spo2, aes(x = spo2_VPO, fill=altitude_cat)) +
  geom_histogram(position = "identity", alpha = 0.4)

qqPlot(spo2_VPO ~ altitude_cat)
```

Distribution deviates from normal and small group size for the moderate altitude group. Will assess non-parametrically.  
```{r}
median_spo2 <- data_spo2 %>% 
  summarize(n = n(),
            spo2_median = median(spo2_VPO),
            Q1 = quantile(spo2_VPO,0.25), 
            Q3 = quantile(spo2_VPO,0.75), 
            min = min(spo2_VPO), 
            max = max(spo2_VPO)
            )
median_spo2
```  

```{r}
wil <- wilcox.test(spo2_VPO ~ altitude_cat, data = data_spo2, exact = FALSE)
wil
```

> The median SpO2 was not different between low and moderate altitude categories (p=`r round(wil$p.value,3)`).  

```{r, include=FALSE}
rm(data_spo2,median_spo2,wil)
```


### SpO2 and hemoglobin      

```{r}
summary(hb)
```


Scatterplot  
```{r}
plot(spo2_VPO~hb, data=data, 
     main="Scatterplot", 
     xlab="Hemoglobin (g/dL)", 
     ylab="SpO2 (%)"
     )
abline(lm(spo2_VPO~altitude),col="red")
```

There does not seem to be a clear pattern.          

Would a smooth term be useful to model SpO2?   
```{r}
ggplot(data, aes(hb,spo2_VPO)) + 
  geom_point(size=0.6,color="gray40") + 
  geom_smooth(method="loess", color="red") +
  ylab("SpO2 (%)") + xlab("Hemoglobin (g/dL)") + 
  ylim(85,100) +
  theme_bw() + 
  theme(panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(colour = "black"),
        axis.text.x = element_text(size=rel(1.2)), 
        axis.text.y = element_text(size=rel(1.2))
        )
```

Hemoglobin likely has an effect on SpO2 at lower hemoglobin values, which makes sense with what is observed in the graph. Assuming a linear relationship could lead to incorrect conclusions according to this. Nonetheless, it looks like the apparent non-linear relationship at low Hb values is due to only 2 observations with wide confidence intervals showing that the true slope could go either up, straight or down, so it may also be incorrect to assume a non-linear relationship based only on this plot. I will model to see if there is an optimal smooth term for hemoglobin or if a linear term best fits the data:    

GAM model with k=2:   
```{r}
model<-gam(spo2_VPO~s(hb,k=2))
summary(model)
k.check(model)
plot(model)
```

GAM model with k=4:   
```{r}
model<-gam(spo2_VPO~s(hb,k=4))
summary(model)
k.check(model)
plot(model)
```

The estimated degrees of freedom (edf) in both cases were 1, plus p=0.6, meaning that a linear term is better fitted to this data than a non-linear term.   

```{r}
spearman <- cor.test(spo2_VPO,hb,method="spearman",exact=FALSE)
spearman
```

> SpO2 and hemoglobin were not correlated (rho= `r round(spearman$estimate,3)`, p=`r round(spearman$p.value,3)`).

```{r, include=FALSE}
rm(spearman,model)
```


### Sex and sleep apnea   

Mean expected frequency:  
```{r}
mean_exp <- data %>% 
  drop_na(sex,sleep_apnea) %>%
  summarize(mean_expected_freq = n()/(nlevels(sex)*nlevels(sleep_apnea)))

mean_exp
```
Since value is grater than 5.0, chi-squared without continuity correction is appropriate.

```{r}
freq <- table(sex, sleep_apnea)
freq
round(prop.table(freq),4)
```

Mosaic Plot  
```{r, fig.height=3, fig.width=5}
ggplot(data = data) +
  geom_mosaic(aes(x = product(sex,sleep_apnea), fill=sex),na.rm = TRUE) +
  scale_fill_manual(values=c("peachpuff","sandybrown")) +
  theme_mosaic() 
```

```{r}
chi <- chisq.test(freq, correct=FALSE)
chi
```

Percentage by level (women, men):  
```{r}
round(prop.table(freq,1),4)*100
```

```{r, include=FALSE}
rm(mean_exp,freq, chi)
```

> Sex was associated with OSA (p<0.001) as men had the diagnosis more frequently compared to women. 

     
