---
title: "Preoperative Atelectasis"
subtitle: "Part 2: Descriptive characteristics and map"
author: "Javier Mancilla Galindo"
date: "`r Sys.Date()`"
execute: 
  echo: true
  warning: false
toc: true
toc_float: true
format:
  html:
    embed-resources: true
  pdf: 
    documentclass: scrartcl
editor: visual
---

# Setup

#### Packages used

```{r}
if (!require("pacman", quietly = TRUE)) {
  install.packages("pacman")
}


pacman::p_load(
  tidyverse, # Used for basic data handling and visualization.
  table1, #Used to create table of descriptive characteristics of sample.
  RColorBrewer, #Color palettes for data visualization. 
  gridExtra, #Used to arrange multiple ggplots in a grid.  
  grid, #Used to arrange multiple ggplots in a grid.
  rnaturalearth, #Used to extract geographical data to create maps. 
  rnaturalearthhires, #Used together with the prior package to create map. 
  sf, #Used together with the prior package to create map. 
  plotly, #Used together with prior two packages to create map. 
  reticulate, #Required together with plotly
  flextable, #Used to export tables.  
  officer,  #Used to export tables.
  report, #Used to cite packages used in this session.   
  webshot2 #Required to render this document as pdf.  
)
```

```{r}
#| include: false

# If you get an error linked to loading rnaturalearthhires, please run
# this code. If not, ignore this chunk of code.   

if (!require("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

remotes::install_github("ropensci/rnaturalearthhires")

pacman::p_load(rnaturalearthhires)
```

##### Session and package dependencies

```{r}
#| echo: false

# Credits chunk of code: Alex Bossers, Utrecht University (a.bossers@uu.nl)

# remove clutter
session <- sessionInfo()
session$BLAS <- NULL
session$LAPACK <- NULL
session$loadedOnly <- NULL
# write log file
writeLines(
  capture.output(print(session, locale = FALSE)),
  paste0("sessions/",lubridate::today(), "_session_Part_2.txt")
)

session
```

```{r}
#| include: false

# Create directories for sub-folders 
figfolder <- "../results/output_figures"
tabfolder <- "../results/output_tables"
psfolder <- "../data/processed"
dir.create(figfolder, showWarnings = FALSE)
dir.create(tabfolder, showWarnings = FALSE)
dir.create(psfolder, showWarnings = FALSE)
```

```{r}
#| include: false

# Load dataset  
data <- read.csv("../data/processed/atelectasis_included.csv",
                 na.strings="NA", 
                 row.names = NULL)
# Recode variables 
source("scripts/variable_names.R")

```

# State of residence of participants

Map generated with the accompanying script ***Map_USA_Canada.R***

This map was built by partly using code adapted from [contribution by cpsievert](https://community.rstudio.com/t/combining-usa-canada-specific-province-with-plotly-plot-geo-or-ggplot2/12375).

```{r}
#| echo: false
#| warning: false

# This may take a few seconds to run, please be patient and ignore warnings. 

source("scripts/Map_USA_Canada.R")
fig

rm(canada, usa, fig, fig_add, figS2, table, us_laea)
```

# Distribution of numerical variables

Distributions were examined with the accompanying sourced function ***distribution_numerical_variables.R***

```{r, fig.height = 4, fig.width = 6}
#| include: false

# If plots are not showing and you wish to see them, 
# please change the above to ## include: true ##  

source("scripts/distribution_plots.R")
distribution_plots(data)
```

Near normal distribution:\
- Age: light tails\
- height: heavy right tail, 4 outliers right\
- hb: heavy tails, bilateral outliers\
- hct: heavy tails, bilateral outliers\
- leu: near normal, bilateral outliers\
- neu_absolute: heavy right tail, two right outliers\
- linf_absolute: heavy right tail, bilateral outliers (more right)\
- mon_absolute: heavy right tail, bilateral outliers (more right)\
- platelets: two right outliers\
- urea: four right outliers\
- creatinine: three right outliers

Distribution not normal:\
- Weight: right-skewed, outliers are verified observations of extreme weight.\
- BMI: right-skewed, outliers are verified observations of extreme BMI.\
- spo2_VPO: Left-skewed\
- neu_percent: left-skewed\
- linf_percent: right-skewed\
- glucose: right-skewed\
- mon_percent: observations around only 5 data points. Will not use this variable, only absolute monocytes will be used.\
- altitude: distribution not clear as values are quite apart an concentrate around single states with diferring mean altitudes. Will attempt to model a smooth term or categorical term in subsequent analyses.

Outcome variable:\
- atelectasis_percent: Zero-inflated. Would be difficult to manage as categorical ordinal due to low number of patients in some categories. Will re-assess alongside subsequent analyses to decide.

# Characteristics of participants

Table 1 generated with the accompanying sourced script ***table1_arguments.R***

```{r}
#| include: false

source("scripts/table1_arguments.R", local = knitr::knit_global())
```

Characteristics of participants are shown for the total sample and by obesity class category as defined by the [World Health Organization](https://www.who.int/europe/news-room/fact-sheets/item/a-healthy-lifestyle---who-recommendations):

-   Class 1, BMI (30-35) kg/m2\
-   Class 2, BMI (35-40) kg/m2\
-   Class 3, BMI \>40 kg/m2

> Characteristics of participants according to BMI class are shown in **Table 1**.

## Table 1

```{r}
#| echo: false

table1
```

NOTE: The **ASA** physical status variable has not been included in analyses since the [updated version of ASA](https://www.asahq.org/standards-and-practice-parameters/statement-on-asa-physical-status-classification-system) consulted in October 2023 classifies obesity (30\<BMI\<40) as ASA 2 and obesity (BMI ≥40) as ASA 3. The distribution of frequencies of ASA\~obesity class in this dataset does not match such definition. This occurred since an outdated version of ASA that did not include obesity was likely used by clinicians when writing the preoperative assessment medical note:

```{r}
#| echo: false

table(data$ASA,data$type_obesity)
```

# Package references

```{r}
#| include: false
report::cite_packages(session)
```

-   Auguie B (2017). *gridExtra: Miscellaneous Functions for "Grid" Graphics*. R package version 2.3, <https://CRAN.R-project.org/package=gridExtra>.
-   Chang W (2023). *webshot2: Take Screenshots of Web Pages*. R package version 0.1.1, <https://CRAN.R-project.org/package=webshot2>.
-   Gohel D, Moog S (2024). *officer: Manipulation of Microsoft Word and PowerPoint Documents*. R package version 0.6.5, <https://CRAN.R-project.org/package=officer>.
-   Gohel D, Skintzos P (2024). *flextable: Functions for Tabular Reporting*. R package version 0.9.5, <https://CRAN.R-project.org/package=flextable>.
-   Grolemund G, Wickham H (2011). “Dates and Times Made Easy with lubridate.” *Journal of Statistical Software*, *40*(3), 1-25. <https://www.jstatsoft.org/v40/i03/>.
-   Makowski D, Lüdecke D, Patil I, Thériault R, Ben-Shachar M, Wiernik B (2023). “Automated Results Reporting as a Practical Tool to Improve Reproducibility and Methodological Best Practices Adoption.” *CRAN*. <https://easystats.github.io/report/>.
-   Massicotte P, South A (2023). *rnaturalearth: World Map Data from Natural Earth*. R package version 1.0.1, <https://CRAN.R-project.org/package=rnaturalearth>.
-   Müller K, Wickham H (2023). *tibble: Simple Data Frames*. R package version 3.2.1, <https://CRAN.R-project.org/package=tibble>.
-   Neuwirth E (2022). *RColorBrewer: ColorBrewer Palettes*. R package version 1.1-3, <https://CRAN.R-project.org/package=RColorBrewer>.
-   Pebesma E, Bivand R (2023). *Spatial Data Science: With applications in R*. Chapman and Hall/CRC. doi:10.1201/9780429459016 <https://doi.org/10.1201/9780429459016>, <https://r-spatial.org/book/>. Pebesma E (2018). “Simple Features for R: Standardized Support for Spatial Vector Data.” *The R Journal*, *10*(1), 439-446. doi:10.32614/RJ-2018-009 <https://doi.org/10.32614/RJ-2018-009>, <https://doi.org/10.32614/RJ-2018-009>.
-   R Core Team (2024). *R: A Language and Environment for Statistical Computing*. R Foundation for Statistical Computing, Vienna, Austria. <https://www.R-project.org/>.
-   Rich B (2023). *table1: Tables of Descriptive Statistics in HTML*. R package version 1.4.3, <https://CRAN.R-project.org/package=table1>.
-   Rinker TW, Kurkiewicz D (2018). *pacman: Package Management for R*. version 0.5.0, <http://github.com/trinker/pacman>.
-   Sievert C (2020). *Interactive Web-Based Data Visualization with R, plotly, and shiny*. Chapman and Hall/CRC. ISBN 9781138331457, <https://plotly-r.com>.
-   South A, Michael S, Massicotte P (????). *rnaturalearthhires: High Resolution World Vector Map Data from Natural Earth used in rnaturalearth*. R package version 1.0.0.9000, https://github.com/ropensci/rnaturalearthhires, <https://docs.ropensci.org/rnaturalearthhires>.
-   Ushey K, Allaire J, Tang Y (2024). *reticulate: Interface to 'Python'*. R package version 1.35.0, <https://CRAN.R-project.org/package=reticulate>.
-   Wickham H (2016). *ggplot2: Elegant Graphics for Data Analysis*. Springer-Verlag New York. ISBN 978-3-319-24277-4, <https://ggplot2.tidyverse.org>.
-   Wickham H (2023). *forcats: Tools for Working with Categorical Variables (Factors)*. R package version 1.0.0, <https://CRAN.R-project.org/package=forcats>.
-   Wickham H (2023). *stringr: Simple, Consistent Wrappers for Common String Operations*. R package version 1.5.1, <https://CRAN.R-project.org/package=stringr>.
-   Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R, Grolemund G, Hayes A, Henry L, Hester J, Kuhn M, Pedersen TL, Miller E, Bache SM, Müller K, Ooms J, Robinson D, Seidel DP, Spinu V, Takahashi K, Vaughan D, Wilke C, Woo K, Yutani H (2019). “Welcome to the tidyverse.” *Journal of Open Source Software*, *4*(43), 1686. doi:10.21105/joss.01686 <https://doi.org/10.21105/joss.01686>.
-   Wickham H, François R, Henry L, Müller K, Vaughan D (2023). *dplyr: A Grammar of Data Manipulation*. R package version 1.1.4, <https://CRAN.R-project.org/package=dplyr>.
-   Wickham H, Henry L (2023). *purrr: Functional Programming Tools*. R package version 1.0.2, <https://CRAN.R-project.org/package=purrr>.
-   Wickham H, Hester J, Bryan J (2024). *readr: Read Rectangular Text Data*. R package version 2.1.5, <https://CRAN.R-project.org/package=readr>.
-   Wickham H, Vaughan D, Girlich M (2024). *tidyr: Tidy Messy Data*. R package version 1.3.1, <https://CRAN.R-project.org/package=tidyr>.

```{r}
#| include: false

# Run this chunk if you wish to clear your environment and unload packages.

rm(properties, name, table1, rndr, normal, nonormal, distribution_plots,
   abbreviations, abbreviations_stats,data, figfolder, psfolder, tabfolder, session)

pacman::p_unload(negate = TRUE)
```
